---
title: "Looping"
author: "Brandon Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Looping}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

#Looping
This package contains a flexible framework for extending the pipe into a loop. The basic idea is this: I often run into the problem of wanting to access an unnamed intermediate in a pipe. Why? A basic strategy of working with data frames is to focus on a certain aspect of the data frame, make some changes, and then reincorporate these changes into the original data frame. This work-flow is best understood through illustration.

Note: this tutorial assumes familiarity with Hadley Wickham's `dplyr` and `magrittr` packages. If you don't know what I'm talking about, go look them up. Your life is about to get a whole lot easier.

##Set-up

Import useful libraries for chaining, and of course, `loopr`.
```{r, message=FALSE}
library(loopr)
library(dplyr)
library(magrittr)
library(knitr)
```

```{r}
loop = loop$new()
```

Set up an extremely simple data frame for illustration.
```{r}
id = c(1, 2, 3, 4)
toFix = c(0, 0, 1, 1)
group = c(1, 1, 1, 0)
example = data_frame(id, toFix, group)
kable(example)
```

`loopr` relies a stack framework.

## Stack
```{r}
stack = stack$new()
````

We can `push` data onto the `stack` like this.
```{r}
stack$push(1)
stack$push(2)
stack$push(3)
````

We can `peek` at the top of the `stack`:
```{r}
stack$peek
````

or at the whole thing.
```{r}
stack$stack
````

We can find the `height` of the `stack` as well:
```{r}
stack$height
```

We can also `pop` off items from the `stack`:
```{r}
stack$pop
stack$pop
stack$pop
```

Now the `stack` is empty.
```{r}
stack$stack
```


##Loop
Why is this important? A `loop` object inherits from a `stack`, and the `begin` method is simply a copy of `push`.

```{r}
"first" %>%
loop$begin()

loop$stack
```

There are two ways to end a loop. You can `end` the loop, or `cross` the loop. Ending the loop takes a function, and uses a `pop` from the `loop` `stack` as the first argument, and its own first argument (or chained argument) as the second.

```{r}
"second" %>%
  loop$end(paste)
```

`cross` is nearly identical, but the order of the arguments gets reversed.

```{r}
"first" %>%
  loop$begin()

"second" %>%
  loop$cross(paste)
```

This is much easier to explain in code than in words.

```{r, eval=FALSE}
end(endData, FUN, ...) = FUN(stack$pop, endData, ...)

cross(crossData, FUN, ...) = FUN(crossData, stack$pop, ...)
```

There are two useful ending functions that are included in this package:`insert` and `amend`. Why are special ending functions needed? In general, traditional join functions are not well suited to the amend work-flow. That is, focusing on a certain aspect of a data frame, making some changes, and then reincorporate these changes into the original data frame. We need `insert` and `amend` to prioritize the information in the amended data frame over the information in the original.

##Insert
`insert` is the slightly more simple case. Let's use our example data again.

Create a set of data to `insert`.
```{r}
insertData =
  example %>%
  filter(toFix == 0) %>%
  mutate(toFix = 1) %>%
  select(-group)
```

Now let's `insert` it back into the original data.
```{r}
insert(example, insertData, by = "id")
```

What happened? Where the `by` variables matched, `insert` excised all rows from `example` and inserted `insertData`. At the end, data was sorted by the `by` variable.

Let's take a look at the slightly more complicated ending function: `amend`.

##Amend
```{r}
amendData = insertData

amend(example, amendData, by = "id")
```

Now the `group` data didn't disappear! If it is not included, `by` defaults to the grouping variables in data.

```{r}
amendData = insertData

example %<>% group_by(id)

amend(example, amendData)
```

A warning: `amend` internally uses the suffix `"toFix"`. If this suffix is already used in your data, modify the `suffix` argument.

`amend` is simply `dplyr::full_join` followed by `amendColumns` to merge non-key columns with matching names. In this case,   `group` from `amendData` overwrites `group` from `example`. How does `amendColumns` work?

###amendColumns
`amendColumns` updates an old set of columns with all non-`NA` values from a matching a new set of columns.

```{r}
oldColumn = c(0, 0)
newColumn = c(1, NA)
data_frame(oldColumn, newColumn) %>%
  amendColumns("oldColumn", "newColumn")
```

###fillColumns
There is also a matching function called  fillColumns . In this function, `NA`'s from `newColumn` are replaced with numbers from the `oldColumn`, but nothing else.

```{r}
oldColumn = c(0, 0)
newColumn = c(1, NA)
data_frame(oldColumn, newColumn) %>%
  fillColumns("newColumn", "oldColumn")
```

##Illustration
Now that we understsand how it works, let's use use our `loop`!

Remind ourselves of what the `example` data looks like.

```{r}
kable(example)
```

Conditional mutation:

```{r}
example %>%
  loop$begin() %>%
    filter(toFix == 0) %>%
    mutate(toFix = 1) %>%
  loop$end(insert, by = "id") %>%
  kable
```

In general, `insert` is best suited to `filter`/`slice` (ungrouped) operations.

Merged `summarize`:

```{r}
example %>%
  group_by(group) %>%
  loop$begin() %>%
    summarize(toFix = mean(toFix)) %>%
  loop$end(amend) %>%
  kable
```

In general, `amend` is best suited to `summarize`/`do` (grouped) operations.

This is only the tip of the iceberg. Do not feel limited to using amend and insert as ending functions. A whole host of others could be useful: join functions, merge functions, even setNames. Loops within loops are in fact quite possible. I would be cautious using them. It can be exhilarating, but make sure to indent each loop carefully.
